A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-22 graph_syntax.tex
finishing the last of the eqn additions

The new syntax is intended to represent three phases of the design problem formulation process. 
In the initial problem definition
  phase, the specific analysis tools and design goals are identified. Next, 
  a single formal problem formulation is identified that specifies design variables, constraints, objectives, 
  analysis tools, and all other elements required to represent the overall MDAO problem. 
Lastly, a specific procedure for solving the problem is selected, e.g. selecting an MDAO optimization architecture. Using the proposed graph syntax, these phases 
  can be represented with the following three graphs:
  \begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem Graph (FPG)
    \item Problem Solution Graph (PSG)
  \end{itemize}

@@

\subsection{Node and Edge Types}
  \label{ss:types}
  We now define different types of nodes and edges with specific properties to compose graphs used to represent an MDAO problem. 
  There are three node types:  
  \begin{description}
    \item[variable:] represents scalar or array data
    \item[model:] responsible for mapping inputs to outputs
    \item[driver:] control logic capable of managing iteration
  \end{description}

@@

  \begin{description}
  \item[connection edge:] Exchange of information between two nodes. These edges 
  can either be fixed (cannot be removed from graph) or free (can be removed). 
  \item [driven edge:] Passing of information from a driver node to a 
    variable node. A single variable node can have many incomming driven edges. These edges are 
    free and can be added or removed from the graph. 
  \end{description}

@@

\subsection{Analysis Blocks and Connections}
\label{ss:analysis blocks and connections}
Analysis tools take in a number input variables and then calculate 
the values for their respective outputs. We represent this process
by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block, each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig.~\ref{f:analysis block}, 
the analysis block includes two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to the outputs of the anlysis. This calculation edge provides the opportunity 
to encode computational cost or other characteristics of the analysis code as 
a weight in a weighted graph formulation. If associating a weight is 
unnecessary, the calculation edge may be omitted and all input and output 
variables can be connected to a single model node for a given analysis block. 
Since all the edges within analysis blocks are fixed, the blocks themselves are 
fixed sub-graphs within the overall MDAO problem graph. The connectivity of 
nodes and edges in an analysis block cannot be altered for the purposes of MDAO 
problem formulation; however, analysis block sub-graphs can be added or removed 
from the MDAO problem graph as needed.

@@

\begin{equation}
  V_{dv} = \{V \st \txt{deg}_l^-(v)=0\}
\end{equation}

@@

The operations of implementing composite objective and constraint functions, 
although typically simple, are effectively the identical to the task performed by an 
analysis block. A composit constraint or objective function can therfore 
be represented as an additional analysis block within the graph, with its own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience, it is useful to distinguish between analysis 
blocks corresponding to analysis codes and those that arise from the addition of 
objectives or constraints to the graph. Therefore, we define an 
\emph{expression block}, as the collection of variable and model nodes related 
to a given objective or constraint function. Fig. \ref{f:obj-cons} highlights 
the objective and constraint blocks from the Sellar problem. If the relevant 
objective and constraint values are direct results from other analysis blocks 
and composite functions are not needed, then expression blocks may be omitted 
from the MDAO problem graph.

@@

  The set of local nodes, $V_l$, is the collection of 
  nodes with exactly one outgoing connection edge.

  \begin{equation}
    V_{l} = \{V \st \txt{deg}^+(v)=1\}
    \label{e:local_definition}
  \end{equation}

@@

  The set of global nodes, $V_g$, is the collection of all nodes with more than 
  one outgoing connection edge.

  \begin{equation}
    V_{g} = \{V \st \txt{deg}^+(v)>1\}
    \label{e:global_definition}
  \end{equation}

@@

  The variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem, $x_1$ is considered a local 
  variable because it directly affects only analysis block $D_1$. By expanding
  $F$ into an expression block with its own inputs, then $F$ is assigned its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global $x_1^*$ 
  node in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the different blocks. Interestingly, the classification of $x_1^*$ 
  as a global variable fits nicely within the structure of the Colaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for setting up 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In that case a global 
  variable is created, and the analysis block again is again assigned an 
  independent local copy of it. Given our definition of global vs local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a given variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  result in the correct structure for CO automatically.

@@

  Looking at the Sellar problem in 
  Fig. \ref{f:sellar_graph_full} we see the expression blocks for each constraint, 
  $G_1$ and $G_2$. Both constraints are local to their respective analysis 
  blocks. But the expression block for $F$ has incomming edges from $D_1$ and 
  $D_2$, so it would considered global.

@@

\subsection{Coupling Between Analyses}

  Coupling exists in a design problem when a set of two or more analyses each depend on the
  other's outputs. In the Sellar problem from Eqn \ref{eqn:simple_fpf} the two 
  coupling constraints, $H_1$ and $H_2$, provide a reciprical dependence between 
  $D_1$ and $D_2$. In a graph, these constraints appear as connection edges between 
  the outputs of each analysis block. The relevant connection edges are highlighted in 
  Fig. \ref{f:coupling}, where they form a closed path between two analysis blocks,
  called a cycle. Cycles of connection edges are the characteristic structure that 
  represents coupling.

@@

  Coupling cycles do not contain driver nodes in the MCG of FPG representation. 
  Solvers, optimizers, and other iterative processes are not invovled in the coupling 
  definition--though one will be necessary to build a valid PSG from a given FPG. 
  Additionally, a coupling cycle has no inherent start or end. It would be acceptable to select
  node in the cycle as a starting point and proceed around the
  loop until you get back to the starting point. For the Sellar Problem, selecting 
  $D_1$ as the starting point would yield a problem as given in 
  Eq.~\ref{eqn:simple}, whereas selecting $D_2$ would yield the problem as given in 
  Eq.~\ref{eqn:simple2}.

@@

  \begin{equation}
    V_{mf} = \{ V \st \txt{deg}^-(v)>1 \} \cup \{ V \st \txt{deg}^-_u(v)>=\txt{deg}^-(v) \}
  \end{equation}

  In a multi-fidelity problem, a given variable node may have multiple incomming 
  edges without causing a collision as defined by by Eq.~\ref{e:collision}. 
  Fig.~\ref{f:collision_example} shows a modified version of the Sellar problem 
  with a new analysis, $D_0$, representing a low fidelity version of $D_1$. 
  In this modified graph, the annotations indicate the upper indegree limit. For 
  $D_2$, $\txt{deg}^-_u(D_2.y_1) = 2$ which avoids a collsion. A  
  $\txt{deg}^-_u(v) > 1$ for any variable in a graph represents a decision to 
  allow multiple fidelities to interact at that part of the graph in order to alieviate
  a conflict. In section \ref{ss:obtaining FPG} we present an algorithm for finding 
  conflicts within a given MCG so that a designer can make the necessary decisisons about each 
  one in turn.

>>>>

  The new syntax is intended to represent three phases of the design problem 
  formulation process. In the initial problem definition phase, the specific 
  analysis tools and design goals are identified. Next, a single formal problem 
  formulation is identified that specifies design variables, constraints, 
  objectives, analysis tools, and all other elements required to represent 
  the overall MDAO problem. Lastly, a specific procedure for solving the problem 
  is selected, e.g. selecting an MDAO optimization architecture. Using 
  the proposed graph syntax, these phases can be represented with the following 
  three graphs:

  \begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem Graph (FPG)
    \item Problem Solution Graph (PSG)
  \end{itemize}

@@

\subsection{Node and Edge Types}
  \label{ss:types}
  We now define different types of nodes and edges with specific properties to compose graphs used to represent an MDAO problem. 
  There are three node types:  
  \begin{description}
    \item[variable:] represents scalar or array data
    \begin{equation} 
      V_{\txt{var}} = \{v \st t_{\txt{node}}(v) = \txt{`variable'}\}
    \end{equation}

    \item[model:] responsible for mapping inputs to outputs
    \begin{equation} 
      V_{\txt{model}} = \{v \st t_{\txt{node}}(v) = \txt{`model'}\}
    \end{equation}

    \item[driver:] control logic capable of managing iteration
    \begin{equation} 
      V_{\txt{driver}} = \{v \st t_{\txt{node}}(v) = \txt{`driver'}\}
    \end{equation}

  \end{description}

@@

  \begin{description}
  \item[connection edge:] Exchange of information between two nodes. These edges 
  can either be fixed (cannot be removed from graph) or free (can be removed). 
    \begin{equation} 
      E_{\txt{conn}} = \{e \st t_{\txt{edge}}(e) = \txt{`connection'}\}
    \end{equation}
  \item [driven edge:] Passing of information from a driver node to a 
    variable node. A single variable node can have many incomming driven edges. These edges are 
    free and can be added or removed from the graph.
     \begin{equation} 
       E_{\txt{driven}} = \{e \st t_{\txt{edge}}(e) = \txt{`driven'}\}
     \end{equation} 
  \end{description}

@@

\subsection{Analysis Blocks and Connections}
\label{ss:analysis blocks and connections}
Analysis tools take in a number input variables and then calculate 
the values for their respective outputs. We represent this process
by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Each analysis block is 
denoted by:

\begin{equation}
  A_i=(V_{A_i},E_{A_i}),\ for\ i\in I, 
\end{equation}

Where $m$ is the number of analysis tools represented in the graph 
and $I=\{1,2,\ldots,m\}$.

Within a given analysis block, variable node represents a single input or 
output and is connected to a single model node via a fixed connection edge. Note that 
in Fig.~\ref{f:analysis block}, the analysis block includes two model nodes, 
with a single edge connecting them. This edge represents the necessary 
calculations to map given inputs to the outputs of the anlysis. This calculation edge provides the opportunity 
to encode computational cost or other characteristics of the analysis code as 
a weight in a weighted graph formulation. If associating a weight is 
unnecessary, the calculation edge and second model node may be omitted and all input and output 
variables can be connected to a single model node for a given analysis block. 
Since all the edges within analysis blocks are fixed, the blocks themselves are 
fixed sub-graphs within the overall MDAO problem graph. The connectivity of 
nodes and edges in an analysis block cannot be altered for the purposes of MDAO 
problem formulation; however, analysis block sub-graphs can be added or removed 
from the MDAO problem graph as needed.

@@

\begin{equation}
  V_{\txt{dv}} = \{V_{\txt{var}} \st \txt{deg}_l^-(v)=0\}
\end{equation}

@@

The operations of implementing composite objective and constraint functions, 
although typically simple, are effectively the identical to the task performed by an 
analysis block. A composit constraint or objective function can therfore 
be represented as an additional analysis block within the graph, with its own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience, it is useful to distinguish between analysis 
blocks corresponding to analysis codes and those that arise from the addition of 
objectives or constraints to the graph. Therefore, we denote an \emph{expression block}, as:

\begin{equation}
  F_j=(V_{F_j},E_{F_j}),\ for\ j\in J, 
\end{equation}

Where $m$ is the number of expressions in the graph
and $J=\{1,2,\ldots,m\}$. For the sake of consistency of naming conventions
$F_j$, $G_j$, and $H_j$ are used to deonte expression blocks for objectives, inequality
constraints, and equality constraints. If the relevant 
objective and constraint values are direct results from other analysis blocks 
and composite functions are not needed, then expression blocks may be omitted 
from the MDAO problem graph. Fig. \ref{f:obj-cons} highlights the 
objective and constraint blocks from the Sellar problem.

@@

  The set of local nodes, $V_\txt{l}$, is the collection of 
  nodes with exactly one outgoing connection edge.

  \begin{equation}
    V_{\txt{l}} = \{V_{\txt{var}} \st \txt{deg}^+(v)=1\}
    \label{e:local_definition}
  \end{equation}

@@

  The set of global nodes, $V_\txt{g}$, is the collection of all nodes with more than 
  one outgoing connection edge.

  \begin{equation}
    V_{\txt{g}} = \{V_{\txt{var}} \st \txt{deg}^+(v)>1\}
    \label{e:global_definition}
  \end{equation}

@@

  The variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem, $x_1$ is considered a local 
  variable because it directly affects only analysis block $A_1$. By expanding
  $F$ into an expression block with its own inputs, then $F$ is assigned its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global $x_1^*$ 
  node in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the different blocks. Interestingly, the classification of $x_1^*$ 
  as a global variable fits nicely within the structure of the Colaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for setting up 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In that case a global 
  variable is created, and the analysis block again is again assigned an 
  independent local copy of it. Given our definition of global vs local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a given variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  result in the correct structure for CO automatically.

@@

  Looking at the Sellar problem in 
  Fig. \ref{f:sellar_graph_full} we see the expression blocks for each constraint, 
  $G_1$ and $G_2$. Both constraints are local to their respective analysis 
  blocks. But the expression block for $F$ has incomming edges from $A_1$ and 
  $A_2$, so it would considered global.

@@

\subsection{Coupling Between Analyses}
  Coupling is described by the presence of a a closed path between two or more
  analysis blocks, called a cycle. A path, given by $P=(V_{path},E_{path})$, 
  contains a cycle if at least two edges in the path refrence the same node, 
  $E = \{(x_0,x_1),(x_1,x_2),\ldots,(x_{k-1},x_k),(x_k,x_0)\}$. These cycles 
  decribe the reciprical dependence of a set of two or more analyses on each 
  other's outputs. In the Sellar problem from Eqn \ref{eqn:simple_fpf} the two 
  coupling constraints, $H_1$ and $H_2$, provide a reciprical dependence between 
  $A_1$ and $A_2$. In a graph, these constraints appear as connection edges between 
  the outputs of each analysis block. The relevant connection edges are highlighted in 
  Fig. \ref{f:coupling}, where they form

@@

  Coupling cycles do not contain driver nodes in the MCG of FPG representation. 
  Solvers, optimizers, and other iterative processes are not invovled in the coupling 
  definition--though one will be necessary to build a valid PSG from a given FPG. 
  Additionally, a coupling cycle has no inherent start or end. It would be acceptable to select
  node in the cycle as a starting point and proceed around the
  loop until you get back to the starting point. For the Sellar Problem, selecting 
  $A_1$ as the starting point would yield a problem as given in 
  Eq.~\ref{eqn:simple}, whereas selecting $A_2$ would yield the problem as given in 
  Eq.~\ref{eqn:simple2}.

@@

  \begin{equation}
    V_{\txt{mf}} = \{ V_{\txt{var}} \st 1 < \txt{deg}^-(v) \leq \txt{deg}^-_u(v) \}
  \end{equation}

  In a multi-fidelity problem, a given variable node may have multiple incomming 
  edges without causing a collision as defined by by Eq.~\ref{e:collision}. 
  Fig.~\ref{f:collision_example} shows a modified version of the Sellar problem 
  with a new analysis, $D_0$, representing a low fidelity version of $A_1$. 
  In this modified graph, the annotations indicate the upper indegree limit. For 
  $A_2$, $\txt{deg}^-_u(A_2.y_1) = 2$ which avoids a collsion. A  
  $\txt{deg}^-_u(v) > 1$ for any variable in a graph represents a decision to 
  allow multiple fidelities to interact at that part of the graph in order to alieviate
  a conflict. In section \ref{ss:obtaining FPG} we present an algorithm for finding 
  conflicts within a given MCG so that a designer can make the necessary decisisons about each 
  one in turn.
